Changes in Faculty Composition and Stagnating State Support for Higher Education

### 2023-03-17 text/sections/00-abstract.tex
Half-way through a major revision.  Needs an intro section on sorting (see Hoxby+ 2008), and revisions of sections 4 (results) onwards.

\noindent
State support for higher education has stagnated on a per-student level, while public universities have reduced their employment of tenure-track or tenured professors, and increased their employment of contingent lecturers (per student).
This analysis relates these changes in the faculty composition at public universities to the fall in per-student university revenues from state appropriations.
A shift-share instrumental variables approach addresses endogeneity in decisions for state appropriations to higher education, by exploiting the yearly change in state support with an institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state appropriations, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Local projection estimates show that these effects linger for up to three years after the initial shock.
Analysis of all the professors at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, implying that these changes in faculty composition arose by reduced hiring of tenure-track and tenured professors at public universities.
These results exhibit the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

\vspace{0.5cm}
\noindent\textbf{Keywords:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labour Markets

\vspace{0.25cm}
\noindent\textbf{JEL Codes:} H72, I23, J45

>>>>

\noindent
State support for higher education per-student has stagnated at the same time as public universities have reduced their employment of tenure-track or tenured professors, and increased their employment of temporary contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per-student revenues from state funding for higher education.
A shift-share instrumental variables approach addresses endogeneity in decisions for state higher education funding by exploiting (1) yearly change in each state's funding for their public universities and (2) each institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state funding, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Local projection estimates show that these effects linger for up to three years after the initial funding shock.
Analysis of all the professors at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, while tenured-track hiring counts fall over this time period, showing that these changes in faculty composition arose by reduced hiring of tenure-track and tenured professors at public universities.
These results exhibit the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

\noindent
\textbf{Key-words:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labour Markets

\noindent
\textbf{JEL Codes:} H72, I23, J45

### 2023-03-17 text/sections/01-introduction.tex
Half-way through a major revision.  Needs an intro section on sorting (see Hoxby+ 2008), and revisions of sections 4 (results) onwards.

\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects, or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding per student, and this affected their composition of faculty.
Falls in public university revenues from the state, instrumented by state-level finance shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, an increase in the number of lecturers, and overall fall in the number of faculty per student.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the wider structure of higher education instruction and research in the US.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.
This fall is driven by stagnating funding provided by state governments, while enrolment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at public universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at public universities, driven by a fall of over 20\% in the ratio of associate and full professors per student at public universities.

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for state funding from shocks to appropriations for the entire state to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in support for higher education and how much each university relies on that support.

Findings show that negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increase in 4.4\% the number of lecturers per student.
Local projection methods show that these effects linger for a small number of years after the initial revenue shock.
Over this time period, state funding fell by around 35\%, while the count of professors per student fell by 9\%, so that these results show that falls in state funding explain around a third of the observed shift away from tenure-track and tenure professors towards contingent faculty.

Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected incumbent professors at public universities.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, yet lecturers' first year salary decreased by 3.8\% in response to a fall in state funding of 10\%.
These findings imply that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures with state appropriation shocks and limits on tuition price.
\cite{NBERw23736} find that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students, but corresponding changes in attendance price (via limits on tuition price) do not have an effect.\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for public university revenues at the state level, combined with rich information on students' individual-level outcomes, to show that increases in state appropriations lead to degree lower completion time and later-life student debt.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

Faculty are a core component of this country's educational and research-innovation university system, yet their composition within universities, and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at private universities.
\cite{abe2015implications} present a model for decision-making by university administrators, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside options.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances and support for higher education.

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
\cite{angrist1999using} first causally show that reducing class size (via religious ruling) induces an increase in test scores for school-age children, yet the magnitude of this effect is not as clear in the higher education setting.
\cite{bandiera2010heterogeneous} use a fixed effect model, combined with multiple individual observations in a long panel, to find that UK university students perform worse academically in particularly large classes, and the difference is largest for students at the top of the academic distribution.
For the composition of faculty teaching university classes, \cite{bettinger2010does,figlio2015tenure} find US students have better enrolment and learning outcomes from courses taught by contingent or adjunct (i.e. not tenured or on tenure-track) professors (relative to tenured or tenure-track).
\cite{ehrenberg2005tenured} observe a negative association between utilisation of non-tenured instructors and student graduation rates.
My approach focuses on count of professors per student at universities, so does not directly observe individual class-sizes, yet presents considers this a possible mechanism through which stagnating state support for higher education negatively affects student outcomes \citep{NBERw23736,NBERw27885}.

>>>>

\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects, or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding per student, and this affected their composition of faculty.
Falls in public university revenues from the state, instrumented by state-level funding shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, and an increase in the number of lecturers.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the wider structure of higher education instruction and research in the US.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.
This fall is driven by stagnating funding provided by state governments, while enrolment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at private universities stayed relatively stable at 0.045, yet fell from 0.04 to around 0.35 professors per student at public universities.
This change was driven by a fall of over 20\% in the count of associate and full professors per student at public universities.

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for state funding, which exploits yearly shocks to the amount an entire state funds higher education to identify the change in a single public unversity's funding.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in support for higher education and how much each university relies on that support.

Negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock.
Over the same time period, state funding fell by around 35\%, while the count of professors per student fell by 9\%, so that these results show that falls in state funding explain around a third of the observed shift away from tenure-track and tenure professors towards contingent faculty.

Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet, the hiring rate of tenure-track professors declined over this time period for all public universities.
This shows that changes in faculty composition came about by hiring reductions of new professors at public universities.

WRITE HERE ABOUT HOW IT IS NOVEL THAT THESE EFFECTS ARE CAUSAL, AND NOT JUST CONCURRENT TRENDS, IN COMPARISON TO HOXBY 2008 SORTING.

A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
and \cite{bound2019public} that these shocks induce public universities to rley more on tuition revenue.
\cite{NBERw27885} use the same methodology to show that increases in state appropriations lead to degree lower completion time\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower higher education completion rate at public universities.
} and later-life debt for students.

Faculty is a core component of the US higher education and research-innovation university system, yet their composition within universities, and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative endowment shocks, and \cite{turner2014impact} for the 2008 recession, while \cite{abe2015implications,johnson2009jep,NBERc13879} model university decision-making and faculty hiring.
Lastly, it has been shown that instruction costs vary substantially across subject areas, primarily thanks to department differences in class size and faculty salaries \citep{hemelt2021math}.

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
\cite{angrist1999using} show that reducing class size (via religious ruling) induces an increase in test scores for school-age children, yet the magnitude of this effect is not as clear in the higher education setting.
\cite{bandiera2010heterogeneous} show that UK university students perform worse academically in particularly large classes.
Regarding faculty composition, the literature is not decided on whether contingent lecturers lead to better (or worse) education outcomes, relative to tenured faculty.
\cite{bettinger2010does,figlio2015tenure} find US students have better enrolment and learning outcomes from courses taught by adjunct professors, while \cite{ehrenberg2005tenured} see that universities who use a greater share of lecturers have lower graduation rates.
\cite{zhu2021limited} note that adjuncts have lower student outcomes than full-time professors, but this effect disappears among individual lecturers when they gain full-time position. 
My approach focuses on count of professors per student at universities, and considers this a possible mechanism through which stagnating state support for higher education negatively affects student outcomes \citep{NBERw23736,NBERw27885}.

### 2023-03-17 text/sections/02-data.tex
Half-way through a major revision.  Needs an intro section on sorting (see Hoxby+ 2008), and revisions of sections 4 (results) onwards.

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information on finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2021 for all relevant variables.
    Figures for enrolment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022},\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrolment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}
so that these institutions are not considered.

@@

\subsection{Trends in Finances, Enrolment, and Faculty Composition}
\label{sec:trends}
A state government plans an annual budget a couple of years ahead of the fiscal year, and the legislature chooses to approve or reject a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
This process leads to yearly variation in state appropriations not seen in other revenues sources (such as federal government appropriations), since US states differ in their support for higher education, and are subject to fiscal constraints.
State appropriations to higher education are a particularly attractive area of state spending to absorb such shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describes the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends.
}
Additionally, there is yearly variation in the number of higher education students in each state \citep{turner2014impact}, so that per-student state appropriations vary on multiple dimensions.

\autoref{fig:funding} exhibits the trends for the mean public university for the years 1990-2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by public universities at a higher rate since.

@@

The state of Illinois experienced the same stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.

@@

The decade of 2010-2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, as is the same for the academic year definition in this analysis. 
    K-12 schools were unaffected thanks to a separate budget agreement allowing non-higher education to operate as before.
}
While state institutions were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring revenues to state agencies and universities for the 2018 fiscal year.
This variation in public university revenues via the state appropriations channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- mean that Illinois public universities exhibit sizable changes in their state appropriations over 2010-2021, of similar order to those for the rest of the country over 1990-2021.

>>>>

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information including finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e., 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990--2017, and manually combine raw NCES data on 2018-2021 for all relevant variables.
    Figures for enrolment, faculty counts and salaries come from the raw NCES version of IPEDS for all years, addressing inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022}, so that these institutions are not considered.

@@

\subsection{Trends in Finances, Enrolment, and Faculty Composition}
\label{sec:trends}

The institutional environment for how states appropriate funding allows for wide variation in state funding for higher education, both between states and across years in each state.
A state government plans an annual budget a couple of years ahead of the fiscal year, and the legislature chooses to approve or reject a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the decision-making process for state appropriations, drawing on administrative records originally analysed by \cite{parmley2009state}.
}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
This process leads to yearly variation in state appropriations not seen in other revenues sources (such as federal government appropriations), since US states differ in their support for higher education, and are subject to fiscal constraints.
State appropriations to higher education are a particularly attractive area of state spending to absorb negative shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describe the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends, and \autoref{fig:funding} observes these same trends.
}
Additionally, there is yearly variation in the number of higher education students in each state \citep{turner2014impact}, so that per-student state appropriations vary on multiple dimensions.

Public university revenues on average stagnated across there years 1990--2021.
\autoref{fig:funding} show the trends in revenues for the mean public university for the years 1990--2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990--2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

@@

There are large differences in the average number of professors per student between the private and public sector.
\autoref{fig:fte-perprof} shows that private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3., with the largest fall in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990, yet have been utilised by public universities at a higher rate since.

@@

The rate that Illinois funds higher education has stagnated to a similar extent as the average for the entire country.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.

@@

The decade of 2010--2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, so is the same as the academic year definition.
}
While public universities were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring funding to state institutions for 2018.
This variation in public university revenues via the state appropriations channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} --- mean that Illinois public universities exhibit sizable changes in their state appropriations over 2010-2021, of similar order to those for the rest of the country over 1990--2021.

### 2023-03-17 text/sections/03-concepts.tex
Half-way through a major revision.  Needs an intro section on sorting (see Hoxby+ 2008), and revisions of sections 4 (results) onwards.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\section{Conceptual Framework}
\label{sec:conceptual}
%
%\textbf{Follow Prof Riehl's comments.
%    Also consider a theory toy-model where reach and teaching are plasuible substitutes, following \cite[Sec.~6.2]{NBERc13879}.
%    Better motivated if research spending is affected.
%}
%The, revenue shock induced, shift away from tenured profs may be mechanically relate to the shift between departments (towards STEM).
%Find a way to mix the “non-replacement of profs” with that for “department composition shift.”
%
%\subsection{University Responses to Financial Shocks}
%\label{sec:responses}

@@

Universities every year hire a number of new professors, either to expand their departments or to replace leaving professors.
The number of professors hired is usually highest among non-tenured adjunct faculty, as these instructors are by majority hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Secondly, tenure-track assistant professors are hired in most years at public universities to replace leaving assistant professors, and are most often hired with a four to six year contract and agreement to be considered for tenure at the end of this term.
Lastly, tenured professors are those that have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors may be hired from outside a university, though this position has the lowest hiring rate among most universities.
It is common for universities to restrict their faculty hiring in the face of financial shocks.
Since the yearly hiring rate is highest among lecturers and assistant professors, financial shocks may lower the number of lecturer and assistant professors at public universities the most, if faculty hiring is affected by falls in state funding.

Additionally, professors' salaries may be disrupted by the stagnation in public university finances.
When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires could be less likely to accept a lower offer from a public university, and we would see that new tenured hires are not affected salary-wise.\footnote{
    \cite{blackaby2005} describes the academic outside options among economist faculty, documenting how they differ for academic position and gender.
}
Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
    Indeed, Cornell University implemented both hiring freezes and even a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the salary cut was returned to professors in the year 2021.
}
Faculty are not paid the same by position, so that their are financial consequences for faculty being promoted between positions within a university.
In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors, or to promote tenured professors from associate to full professor either.

It is not immediately clear which effect will dominate, and which position of professor will be effected the most by changes in their university's finances.
Yet, there is one empirical fact worth noting: a mean assistant or tenured professor earns more than double that of an adjunct lecturer per year (\autoref{tab:illinois-summary}), and the tenure contract means universities are compelled to pay that higher salary for more years than contingent lecturers.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers to achieve this obligation.\footnote{
    That is substituted away from tenure-track and tenured professors according to the salary ratio, and the relative productivity between these two positions.
    It is not possible to estimate the relative productivity in this paper, so I leave the rate of substitution between the two positions to further research.
}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\section{Conceptual Framework}
\label{sec:conceptual}

@@

Most universities hire a number of new professors every year, either to expand their departments or to replace leaving professors.
The number of professors hired in each year is usually highest among non-tenured adjunct faculty, as these instructors are mostly hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Secondly, tenure-track assistant professors are hired to replace leaving assistant professors, and are most often hired with a four to six year contract and agreement for tenure consideration.
Lastly, tenured professors have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors may be hired from outside a university, though this position has the lowest yearly hiring rate among most universities.
It is common for universities to restrict their faculty hiring in the face of financial shocks.
Since the yearly hiring rate is highest among lecturers and assistant professors, financial shocks may lower the number of lecturer and assistant professors at public universities the most, if faculty hiring is affected by falls in funding.

Additionally, faculty salaries may by affected by changes in public university finances.
When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires may be less likely to accept a lower salary offer from a public university.\footnote{
    Faculty often consider their outside options, as in offers for employment and/or promotion from other universities.
    See \cite{blackaby2005} for an overview of faculty outside options.
}
Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
    Indeed, Cornell University implemented hiring freezes and a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the cut was returned to professors in the year 2021.
}
Faculty are not paid the same by position, so that there are financial consequences for a university promoting their own faculty.
In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors (as tenured professors generally receive a salary raise and long term contract), or to promote tenured professors from associate to full professor.

It is not immediately clear which effects will dominate, or whether tenured professors vs lecturers will be more affected by changes in their university's finances.
Yet, there is one empirical fact worth noting: a mean assistant or tenured professor earns more than double that of an adjunct lecturer per year (see \autoref{tab:illinois-summary}), and the tenure contract means universities must pay a higher salary for a guaranteed number years, more so than for contingent lecturers.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers to achieve this obligation.\footnote{
    That is, substituted away from tenure-track and tenured professors according to the salary ratio, and the relative productivity between these two positions.
}
